---
title: "STAT 331 Final Portfolio Reflection"
author: "Marietta Nikolskaya"
format: html
execute:
  eval: false
---

# Grade Earned

I believe I have earned a grade of **B+** in STAT 331.

---

# Learning Target Evidence

## Working with Data

### WD-1: Import data from various formats
```{r}
# Lab 3 - read_csv()
teacher_evals <- read_csv("teacher_evals.csv")
```
**Explanation:** This demonstrates my ability to import CSV files using `read_csv()` from the readr package, which is part of the tidyverse.

### WD-2: Select necessary columns
```{r}
# Lab 3 Question 5
teacher_evals_clean <- teacher_evals |> 
  select(course_id, teacher_id, question_no, no_participants, 
         resp_share, SET_score_avg, percent_failed_cur, 
         academic_degree, seniority, sex)
```
**Explanation:** I used `select()` to keep only the relevant columns needed for analysis, removing unnecessary variables to streamline the dataset.

### WD-3: Filter rows for various data types
```{r}
# Lab 5 - Filtering with multiple data types
crime_scene_report <- crime_scene_report |>
  filter(date == 20180115,          # numeric
         type == "murder",           # character
         city == "SQL City")         # character

witness_2 <- person |>
  filter(str_detect(name, "Annabel"),  # character with pattern
         address_street_name == "Franklin Ave")
```
**Explanation:** This code shows my proficiency with filtering different data types including numeric dates, character strings, and using string pattern detection with `str_detect()`.

### WD-4: Modify existing and create new variables
```{r}
# Lab 3 Question 5 - Type conversion and renaming
teacher_evals_clean <- teacher_evals |> 
  rename(sex = gender) |> 
  mutate(teacher_id = as.character(teacher_id), 
         course_id = as.character(course_id),
         question_no = as.integer(question_no),
         academic_degree = as.factor(academic_degree),
         sex = as.factor(sex))
```
**Explanation:** I modified multiple variables by converting them to appropriate data types (character, integer, factor) to prepare the data for analysis.

### WD-5: Use mutating joins ⭐ (Required for B)
```{r}
# Lab 5 - Complex three-way join
final_suspect <- gym_jan9_suspect %>%
  inner_join(person, by = c("person_id" = "id")) %>%
  inner_join(car_suspect, by = c("license_id" = "id"))
```
**Explanation:** This demonstrates sophisticated use of `inner_join()` to combine three datasets (gym members, person records, and driver's licenses) using different join keys to solve the murder mystery.
```{r}
# Lab 4 Question 3 - Join with renamed columns
ca_childcare <- ca_childcare |> 
  left_join(tax_rev, by = c(
    "county_name" = "entity_name", 
    "study_year" = "year"
  ))
```
**Explanation:** I used `left_join()` with renamed columns to merge tax revenue data with childcare data, matching on county name and year.

### WD-7: Pivot datasets ⭐ (Required for A)
```{r}
# Lab 4 Question 5 - pivot_wider()
ca_median_income <- ca_childcare |> 
  filter(study_year %in% c(2008, 2018)) |> 
  group_by(census_region, study_year) |>
  summarize(median_income = median(mhi_2018)) |>
  pivot_wider(names_from = study_year, 
              values_from = median_income) |> 
  arrange(desc('2018'))
```
**Explanation:** I pivoted data from long to wide format, creating separate columns for 2008 and 2018 median incomes by region.
```{r}
# Lab 4 Question 7 - pivot_longer()
data_plot <- ca_childcare |> 
  pivot_longer(
    cols = c(mc_infant, mc_toddler, mc_preschool),
    names_to = "age_group",
    values_to = "price"
  )
```
**Explanation:** I pivoted data from wide to long format to prepare it for visualization, combining three age group columns into one variable.

---

## Reproducibility

### R-1: Create reproducible analyses
```{r}
# Lab 2 - YAML header with execution options
---
title: "Lab 2: Exploring Rodents"
author: "Marietta Nikolskaya"
format: html
execute:
  warning: false
  message: false
  echo: true
---
```
**Explanation:** All my lab assignments used proper YAML headers, code chunk options, and organized structure to create professional, reproducible documents.

### R-2: Write well-documented and tidy code
```{r}
# Lab 5 - Well-commented code
# Finding witness 1 - lives at last house on Northwestern Dr
witness_1 <- person %>% 
  filter(address_street_name == "Northwestern Dr") %>%  
  slice_max(address_number)

# Finding witness 2 - named Annabel on Franklin Ave
witness_2 <- person %>%
  filter(str_detect(name, "Annabel"),
         address_street_name == "Franklin Ave")
```
**Explanation:** My code throughout the quarter included clear comments explaining each step, used proper spacing and indentation, and followed tidyverse style conventions.

---

## Data Visualization & Summarization

### DVS-1: Create visualizations for various variable types
```{r}
# Lab 2 - Scatterplot with numeric variables
ggplot(data = surveys, 
       aes(x = weight, y = hindfoot_length)) + 
  geom_point(alpha = 0.4) + 
  facet_wrap(~ species)
```
**Explanation:** I created visualizations for numeric continuous variables (weight, hindfoot length) using scatterplots with appropriate transparency to handle overplotting.

### DVS-2: Use plot modifications ⭐ (Required for B)
```{r}
# Challenge 3 - Stacked bar plot with custom styling
ggplot(teacher_evals_compare, aes(x = sen_level, fill = SET_level)) +
  geom_bar(position = "fill") + 
  geom_text(
    aes(label = after_stat(count)),
    stat = "count",
    position = position_fill(vjust = 0.5),
    color = "white",
    size = 3.5
  ) +
  labs(
    title = "Evaluation of Teachers' Use of Activities",
    x = "Years of Experience",
    y = NULL,  
    fill = "Evaluation Rating"
  ) +
  scale_fill_manual(
    values = c("excellent" = "#B19CD9", "standard" = "#C8A882")
  ) +
  theme_minimal() + 
  theme(legend.position = "top")
```
**Explanation:** This plot demonstrates extensive modifications including custom colors, text labels with positioning, clear titles and labels, and theme customization to make the visualization clear and professional.

### DVS-3: Show creativity in visualizations ⭐ (Required for A)
```{r}
# Challenge 2 - Embedding legend in plot title using ggtext
library(ggtext)

ggplot(surveys, aes(x = weight, y = species, fill = sex)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(color = "grey45", alpha = 0.35, width = 0, height = 0.1) +
  labs(
    title = "Distribution of Body Weight by Species",
    subtitle = paste0(
      "Sex: ",
      "<span style='color:#4C78A8;'>Male</span> • ",
      "<span style='color:#E4572E;'>Female</span>"
    ),
    x = "Body Weight (grams)",
    y = "Species"
  ) +
  scale_fill_manual(values = c("F" = "#E4572E", "M" = "#4C78A8")) +
  theme_bw() +
  theme(
    legend.position = "none",
    plot.subtitle = element_markdown(size = 12),
    panel.grid.minor = element_blank()
  )
```
**Explanation:** This demonstrates creativity by embedding the legend directly into the plot subtitle using HTML styling and the ggtext package, eliminating the need for a separate legend and making the plot more elegant and easier to read.

### DVS-4: Calculate numerical summaries
```{r}
# Lab 3 Question 7 - Multiple summary statistics
teacher_evals_clean |> 
  distinct(teacher_id, .keep_all = TRUE) |>
  summarize(
    min_seniority = min(seniority),
    max_seniority = max(seniority),
    mean_seniority = mean(seniority),
    median_seniority = median(seniority)
  )
```
**Explanation:** I calculated multiple numerical summaries (min, max, mean, median) to fully describe the distribution of instructor seniority.

### DVS-5: Calculate summaries across groups ⭐ (Required for B)
```{r}
# Lab 3 Question 10
teacher_evals_clean |>
  filter(question_no == 901) |>
  group_by(teacher_id) |>
  summarize(
    avg_rating = mean(SET_score_avg),  
    num_courses = n_distinct(course_id)  
  ) |>
  filter(num_courses >= 5) |>
  filter(
    avg_rating == min(avg_rating) |
    avg_rating == max(avg_rating)
  )
```
**Explanation:** I grouped data by teacher_id and calculated average ratings and course counts for each instructor, then filtered to find the highest and lowest performers among those teaching at least 5 courses.

### DVS-6: Create clear tables ⭐ (Required for B)
```{r}
# Lab 4 Question 5 - Clean summary table
ca_median_income <- ca_childcare |> 
  filter(study_year %in% c(2008, 2018)) |> 
  group_by(census_region, study_year) |>
  summarize(median_income = median(mhi_2018)) |>
  pivot_wider(names_from = study_year, 
              values_from = median_income) |> 
  arrange(desc('2018'))
```
**Explanation:** This table is clearly structured with regions in rows, years in columns, and sorted by 2018 income for easy comparison across time and regions.

---

## Program Efficiency

### PE-1: Write concise code without repetition ⭐ (Required for B)
```{r}
# Lab 5 - Efficient pipeline without repetition
mastermind <- woman_names %>% 
  inner_join(concert, by = c("id.y" = "person_id"))
```
**Explanation:** Instead of writing separate filter and merge operations, I used efficient pipelines that combined operations, avoiding code repetition throughout my assignments.

### PE-4: Use modern tools ⭐ (Required for B)
```{r}
# Lab 3 Question 8 - Using if_any() for missing values
teacher_evals_clean |>
  filter(if_any(
    .cols = everything(), 
    .fns = ~ is.na(.x)
  ))
```
**Explanation:** I used modern tidyverse functions like `if_any()` with formula syntax instead of older base R approaches, demonstrating knowledge of current best practices.
```{r}
# Lab 3 - Using n_distinct() for unique counts
teacher_evals_clean |> 
  summarize(
    unique_teachers = n_distinct(teacher_id),
    unique_courses = n_distinct(course_id)
  )
```
**Explanation:** I used `n_distinct()` instead of combining `length()` and `unique()`, showing familiarity with efficient modern tidyverse functions.

---

## Data Simulation & Statistical Models

### DSSM-2: Conduct statistical analyses ⭐ (Required for B)
```{r}
# Challenge 3 - Chi-square test
chisq.test(teacher_evals_compare$SET_level, 
           teacher_evals_compare$sen_level)

# Result: X-squared = 10.207, df = 2, p-value = 0.006075
```
**Explanation:** I conducted a chi-square test of independence and properly interpreted the results, concluding that evaluation level and instructor seniority are not independent (p = 0.006 < 0.05).
```{r}
# Lab 4 Question 9 - Linear regression
reg_mod1 <- lm(mc_infant ~ mhi_2018, data = ca_childcare)
summary(reg_mod1)
```
**Explanation:** I fit a linear regression model to examine the relationship between median household income and infant childcare costs, then interpreted the output including the coefficient, p-value, and R-squared value.
```{r}
# Lab 2 - ANOVA
species_mod <- aov(weight ~ species, data = surveys)
summary(species_mod)
```
**Explanation:** I performed a one-way ANOVA to compare mean weights across 14 rodent species and drew appropriate conclusions based on the F-statistic and p-value.

---

# How I Met the Grade Criteria for B+

## Learning Targets

I have demonstrated **proficiency with all D-level targets** and **most B-level targets**:

**D-level (all mastered):**
- WD-1 through WD-4: Data import, selection, filtering, and manipulation ✅
- R-1 and R-2: Reproducible, well-documented code ✅
- DVS-4 and DVS-5: Numerical summaries and grouped summaries ✅

**B-level (mastered):**
- **WD-5**: Mutating joins demonstrated in Lab 5 (three-way join) and Lab 4 ✅
- **WD-7**: Pivoting demonstrated in Lab 4 with both `pivot_wider()` and `pivot_longer()` ✅
- **DVS-2**: Plot modifications shown extensively in Challenge 3 and Lab 4 ✅
- **DVS-6**: Clear tables demonstrated in Lab 4 and Challenge 4 ✅
- **PE-1**: Concise code throughout all assignments ✅
- **PE-4**: Modern tools like `n_distinct()`, `if_any()`, and lubridate functions ✅
- **DSSM-2**: Statistical analyses including chi-square, linear regression, and ANOVA ✅

**B-level (gaps):**
- **PE-2**: Writing custom functions - *I do not have strong evidence of this target*
- **DSSM-1**: Data simulation - *I do not have strong evidence of this target*

**A-level (partial evidence):**
- **DVS-3**: Creative visualizations shown in Challenge 2 with embedded legend ✅
- **WD-7**: Pivoting (also an A requirement) ✅
- **PE-3**: Iteration - *No evidence*
- **DVS-7**: Creative tables - *No evidence*

## Revising Thinking

I submitted revisions for **more than half** of my assignments throughout the quarter. My revision process included:

- **Lab 3 revisions**: Fixed issues with code formatting (keeping lines under 80 characters), improved use of functions like `if_any()` with proper syntax, and enhanced data descriptions with context about when/where/how data were collected
- **Challenge 3 revisions**: Added percentage labels to my stacked bar plot, improved statistical interpretation to include context about what Question 3 measured, and removed inappropriate use of headers for question responses
- **Reflections included**: For each revision, I documented what I did wrong, what I learned, and why my new approach was better

My reflections demonstrated understanding of not just *what* to fix, but *why* the changes improved my code. I learned from feedback and avoided repeating the same mistakes in future assignments.

## Extending Thinking

I **occasionally** pushed myself beyond the basic requirements:

- **Challenge 2**: I completed the "Hot" option (most difficult), embedding the legend in the plot title using HTML and the ggtext package - going beyond standard ggplot skills
- **Lab 4 Challenge**: I created a lollipop plot in addition to the required visualizations to better show cost differences
- **Challenge 3**: Added custom text labels with `after_stat(count)` to make my stacked bar plot more informative

While I attempted more difficult challenges occasionally, I recognize I could have pushed myself more consistently throughout the quarter.

## Collaborative Group Member

Throughout the quarter, I was **patient and respectful** when collaborating with partners during Practice Activities. I:

- Followed the collaborative protocol by switching between Typer and Talker roles as instructed
- Communicated clearly with my partners about approach and reasoning
- Was open to different ideas and approaches to solving problems
- **However**, I acknowledge I sometimes performed tasks outside my assigned role rather than strictly adhering to the Typer/Talker structure

My collaboration was generally good, but I recognize there's room for improvement in following protocol more strictly.

## Peer-Code Review

I provided peer code reviews **most weeks**, offering:

- **Specific feedback** on code formatting, function usage, and documentation
- **Constructive suggestions** for improvement that were honest and kind
- **Demonstrations** of alternative approaches when appropriate

My reviews showed genuine desire to help classmates improve and grow as data scientists.

---

# Supporting Artifacts in This Portfolio

1. **Lab 5 (Murder in SQL City)** - Demonstrates WD-5 (joins), WD-3 (filtering), R-2 (documentation)
2. **Lab 3 (Student Evaluations - REVISED)** - Demonstrates WD-1 through WD-4, DVS-4, DVS-5, PE-1, PE-4
3. **Challenge 3 (Chi-Square Analysis - REVISED)** - Demonstrates DVS-2 (plot modifications), DSSM-2 (statistical analysis)
4. **Lab 4 (Childcare Costs)** - Demonstrates WD-5 (joins), WD-7 (pivoting), DVS-2, DSSM-2
5. **Lab 4 Challenge** - Demonstrates DVS-3 (creative lollipop plot)
6. **Lab 2 (Rodents)** - Demonstrates DVS-1, DVS-2, DSSM-2
7. **Challenge 2 (ggplot2 Spice - REVISED)** - Demonstrates DVS-3 (creative visualization with embedded legend)

---

# Justification for B+ Rather Than B

While I am missing evidence for two B-level targets (PE-2: functions and DSSM-1: simulation), I believe my work merits a B+ because:

1. **I exceeded B requirements in some areas**: My proficiency with WD-7 (pivoting) and DVS-3 (creative visualizations) are A-level targets that I have demonstrated well

2. **Quality of work**: My revised assignments show deep understanding, with thoughtful reflections on my learning process

3. **Strong performance on most B targets**: I demonstrated proficiency in 7 out of 9 B-level targets, with particularly strong evidence for joins, pivoting, plot modifications, and statistical analyses

4. **Revision effort**: I submitted thoughtful revisions with genuine reflection on my growth, meeting the B-level criterion of "submitting revisions for most assignments with reflections"

I recognize that without evidence of writing functions (PE-2) and data simulation (DSSM-1), I do not fully meet all B criteria. However, my strengths in other areas, including some A-level targets, support a B+ grade as representing my overall achievement in the course.
