---
title: "STAT 331 Final Portfolio Reflection"
author: "Your Name"
format: html
execute:
  eval: false
---

# Grade Earned

I believe I have earned a grade of **[STATE YOUR GRADE HERE: A, A-, B+, B, etc.]** in STAT 331.

---

# Learning Target Evidence

## Learning Target 1: [Target Name]

**Code Example:**
```{r}
# [Assignment Name - e.g., Lab 2 Question 1]

# Paste your code here
```

**Explanation:** [Explain what this code demonstrates about your learning]

---

## Learning Target 2: [Target Name]

**Code Example:**
```{r}
# [Assignment Name]

# Paste your code here
```

**Explanation:** [Explain what this code demonstrates]

---

[Continue for all learning targets...]

---

# How I Met the Grade Criteria

Based on the course syllabus, the criteria for a [YOUR GRADE] include:

1. **[Criterion 1 from syllabus]:** I met this by... [provide specific evidence and reference artifacts]

2. **[Criterion 2 from syllabus]:** I demonstrated this through... [provide specific evidence]

3. **[Continue for all criteria]**

---

# Additional Considerations

[Optional: Add any other factors that support your grade request]

---

# Supporting Artifacts

The following assignments in my `artifacts` folder support this reflection:

- [artifact1.qmd] - demonstrates [what it shows]
- [artifact2.qmd] - demonstrates [what it shows]
- [Continue listing...]
