---
title: "Challenge 7: Incorporating Multiple Inputs"
subtitle: "Functions + Fish"
author: "Marietta Nikolskaya"
format: 
  html:
    code-fold: true
    code-tools: true
    warning: false
    embed-resources: true
editor: source
---

```{r}
#| label: setup
library(tidyverse)
fish <- read_csv("BlackfootFish.csv")

```

## Condition Index

A frequently used measurement for fish health is a condition index [(Wikipedia article)](https://en.wikipedia.org/wiki/Condition_index_in_fish). The following
simple equation can be used to calculate the approximate condition index of a
fish:

$$\text{condition index} = \frac{weight}{length^3} \times 100$$

**1. There are specific units required for the calculation of a condition index -- length must be in centimeters and weight must be in grams. The weight data for the Blackfoot River fish were collected in grams; the length data were collected in millimeters. Transform the length data to the correct units.**

```{r}
#| label: transform-data-to-have-correct-units
fish <- fish |>
  mutate(length = length / 10)
head(fish)

```

**2. Collecting data of this sort can be very messy! Write a function that will replace unlikely `length` and `weight` measurements with `NA`. Your function should accept at least three inputs:**

+ **a vector of measurements,**
+ **the minimum reasonable value,**
+ **the maximum reasonable value.**

**If a value falls outside these bounds, you should replace it with an `NA`.**

**Hint:** If you are struggling with the structure of your function, I would suggest reading the [Mutating Function](https://r4ds.hadley.nz/functions.html#mutate-functions) from R4DS.

```{r}
#| label: function-to-change-impossible-measurements-to-NA

replace_outliers <- function(measurements, min_value, max_value) {

  if (!is.numeric(measurements)) {
    stop("measurements must be numeric")
  }
  
  if (!is.numeric(min_value) | !is.numeric(max_value)) {
    stop("min_value and max_value must be numeric")
  }
  
  if (length(min_value) != 1 | length(max_value) != 1) {
    stop("min_value and max_value must be single values")
  }
  
  if (min_value >= max_value) {
    stop("min_value must be less than max_value")
  }
  
  case_when(
    measurements < min_value ~ NA_real_,
    measurements > max_value ~ NA_real_,
    TRUE ~ measurements
  )
}

```

## Revision: 

**What I did wrong: **
No input validation to check if inputs are the correct type.
Used if_else() with multiple conditions (though this actually works fine here).
Used return() when R automatically returns the last expression.

**What I learned: **
Functions should validate inputs to ensure they're the correct type and make sense.
case_when() is built for multiple conditions and is more readable than complex if_else().
R automatically returns the last expression in a function - no need for return() unless returning early.
Good function checks improve code reliability and provide helpful error messages.

**3. After consulting the [Montana Record Table](https://fwp.mt.gov/fish/anglingData/records) for the four species of trout included in these data, I have conjectured that it is unlikely to have measurements for fish below 5 cm and above 80 cm in length or below 10 g and above 4,000 g in weight. Use your function to modify the `length` and `weight` columns of the `fish` dataset based on my cutoffs.**

```{r}
#| label: modify-impossible-values-of-length-weight
fish <- fish |>
  mutate(
    length = replace_outliers(length, min_value = 5, max_value = 80),
    weight = replace_outliers(weight, min_value = 10, max_value = 4000)
  )
summary(fish$length)
summary(fish$weight)
head(fish)
```

**4. Write a function to calculate the condition index of a fish, given inputs of weight and length.**

```{r}
#| label: condition-function
calculate_condition_index <- function(weight, length) {
  if (!is.numeric(weight)) {
    stop("weight must be numeric")
  }
  
  if (!is.numeric(length)) {
    stop("length must be numeric")
  }
  
  if (length(weight) != length(length)) {
    stop("weight and length must have the same length")
  }
  
  (weight / length ^ 3) * 100
}


fish <- fish |>
  mutate(condition_index = calculate_condition_index(weight, length))
summary(fish$condition_index)

```

## Revision: 

**What I did wrong: **
No input validation.
Unnecessary return() statement.

**What I learned: **
Even mathematical functions need input validation to ensure correct types.
When functions take multiple vectors, check they're compatible lengths.
Removing unnecessary return() makes code cleaner and more R-idiomatic.
The last expression is automatically returned, making the function more concise.


**5. Make a thoughtful visualization of how fish conditions have varied over the duration of this study.** To be thoughtful...you need to use at least a third variable!


```{r}
#| label: condition-indices-over-time

# Create summary data with better ordering
fish_summary <- fish |>
  drop_na(condition_index) |>
  group_by(year, species) |>
  summarise(
    mean_condition = mean(condition_index, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  filter(n >= 5) |>
  mutate(species = fct_reorder2(species, year, mean_condition))

ggplot(fish_summary, aes(x = year, y = mean_condition, color = species)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Mean Fish Condition Index Over Time by Species",
    subtitle = "Yearly averages for groups with n ≥ 5 observations",
    x = "Year",
    y = "Mean Condition Index",
    color = "Species",
    caption = "Point size represents sample size ranging from 5 to several hundred fish per year"
  ) +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_color_brewer(palette = "Set2")

```

## Revision: 

**What I did wrong: **
Legend at bottom makes it harder to match lines to species.
Legend order doesn't match line order on plot.
Sample size as point size is cluttered and hard to read.

**What I learned: ** 
Legend on the right is easier to match with line positions than bottom.
fct_reorder2() can reorder legend items to match the visual order of lines.
Multiple visual encodings (color + size) can create clutter.


