---
title: "Lab 7: Functions + Fish"
author: "Marietta Nikolskaya"
format: 
  html:
    toc: true
    code-fold: true
    embed-resources: true
    code-tools: true
editor: source
execute: 
  warning: false
  
---

The goal of this lab is learn more about exploring missing data and writing
modular code.

```{r}
#| label: setup
library(dplyr)
library(tidyverse)
fish <- read.csv("BlackfootFish.csv")
glimpse(fish)


```

## The Data

This lab's data concerns mark-recapture data on four species of trout from the
Blackfoot River outside of Helena, Montana. These four species are
**rainbow trout (RBT)**, **westslope cutthroat trout (WCT)**, **bull trout**,
and **brown trout**.

Mark-recapture is a common method used by ecologists to estimate a population's
size when it is impossible to conduct a census (count every animal). This method
works by *tagging* animals with a tracking device so that scientists can track
their movement and presence.

## Data Exploration

The measurements of each captured fish were taken by a biologist on a raft in
the river. The lack of a laboratory setting opens the door to the possibility of
measurement errors.

**1. Let's look for missing values in the dataset. Output ONE table that answers BOTH of the following questions:**

+ **How many observations have missing values?**
+ **What variable(s) have missing values present?**

::: callout-tip
# You should use `across()`!
:::

```{r}
#| label: find-missing-values

missing_summary <- fish %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), 
               names_to = "Variable", 
               values_to = "Missing_Count")

missing_summary
```

**2. Use a pivot to transform your summary table from Question 1 into a table that is easier to read.** 

```{r}
#| label: pivot-summary-table

missing_summary %>%
  pivot_wider(names_from = Variable, 
              values_from = Missing_Count)
missing_summary

```
## Revision: 

**What I did wrong:** 
I filtered to only missing values and then selected only one column, hiding the full pivoted structure. 

**What I learned: **
When asked to create a pivot table, I should display the entire transformed structure, not just extract a single summary value. The pivot_wider transformation itself makes the data easier to read by putting variables as columns. 

**3. Create ONE thoughtful visualization that explores the frequency of missing values across the different years, sections, and trips.**


```{r}
#| label: visualize-missing-patterns

fish_missing_viz <- fish %>%
  mutate(
    has_missing = is.na(weight),
    trip_label = paste("Trip", trip)  
  ) %>%
  group_by(year, section, trip_label, has_missing) %>%
  summarise(count = n(), .groups = "drop") %>%
  filter(has_missing == TRUE)  

ggplot(fish_missing_viz, 
       aes(x = as.factor(year), y = count, fill = section)) +
  geom_col(position = "dodge") +
  facet_wrap(~trip_label) +
  labs(
    title = "Missing Weight Measurements Across Mark-Recapture Surveys of Blackfoot River Trout",
    subtitle = "All missing values occur in marked (recaptured) fish across years and river sections",
    x = "Year",
    y = "Number of Missing Weight Values",
    fill = "River Section"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

```

## Revisions: 

**What I did wrong: **
Used mark_status label which has biological meaning separate from missingness
Didn't include trip as requested
Title could provide more context

**What I learned: **
I should include ALL requested variables (year, section, AND trip)
Use fct_recode() or paste() to create informative facet labels like "Trip 1" instead of just "1"
Labels should accurately reflect what's being shown (missingness, not biological status)
Titles and subtitles can provide important biological context about the data collection process

## Rescaling the Data

If I wanted to rescale every quantitative variable in my dataset so that they
only have values between 0 and 1, I could use this code:

```{r}
#| echo: true
#| eval: false

fish <- fish |> 
  mutate(length = (length - min(length, na.rm = TRUE)) / 
           (max(length, na.rm = TRUE) - min(length, na.rm = TRUE)), 
         weight = (weight - min(weight, na.rm = TRUE)) / 
           (max(weight, na.rm = TRUE) - min(length, na.rm = TRUE)))
```


**4. Transform the repeated process above into a `rescale_01()` function. Your function should...**

+ **... take a single vector as input.**
+ **... return the rescaled vector.**

```{r}
#| label: write-rescale-function


rescale_01 <- function(x) {
  if (!is.numeric(x)) {
    stop()
  }
  
  if (length(x) <= 1) {
    stop()
  }
  
  min_x <- min(x, na.rm = TRUE)
  max_x <- max(x, na.rm = TRUE)
  

  (x - min_x) / (max_x - min_x)
}

```

::: callout-tip
# Efficiency 

Think about the efficiency of the function you wrote. Are you calling the
**same** function multiple times? You might want to look into the `range()` 
function. 
:::

**5. Let's incorporate some input validation into your function. Modify your previous code so that the function stops if ...**

+ **... the input vector is not numeric.**
+ **... the length of the input vector is not greater than 1.**

::: callout-tip
# Modify Previous Code

Do not create a new code chunk here -- simply add these stops to your function
above!
:::

## Test Your Function

**6. Run the code below to test your function. Verify that the maximum of your rescaled vector is 1 and the minimum is 0!**

```{r}
#| label: verify-rescale-function

x <- c(1:25, NA)

rescaled <- rescale_01(x)
min(rescaled, na.rm = TRUE)
max(rescaled, na.rm = TRUE)
```

Next, let's test the function on the `length` column of the `BlackfootFish` data.

**7. The code below makes a histogram of the original values of `length`. Add a plot of the rescaled values of `length`. Output your plots side-by-side, so the reader can confirm the only aspect that has changed is the scale.**

::: callout-warning
This will require you to call your `rescale_01()` function within a `mutate()`
statement in order to create a `length_scaled` variable.
:::

```{r}
#| label: compare-original-with-rescaled-lengths

fish |>  
  ggplot(aes(x = length)) + 
  geom_histogram(binwidth = 45) +
  labs(x = "Original Values of Fish Length (mm)",
       y = "Count") +
  scale_y_continuous(limits = c(0, 4000)) +
  theme_minimal()

# Rescaled length plot
fish %>%
  mutate(length_scaled = rescale_01(length)) %>%
  ggplot(aes(x = length_scaled)) + 
  geom_histogram(binwidth = 0.1) +
  labs(x = "Rescaled Values of Fish Length (0-1)",
       y = "Count") +
  scale_y_continuous(limits = c(0, 4000)) +
  theme_minimal()

```

## Revisions: 

**What I did wrong: **
Created two separate code chunks that output plots sequentially rather than side-by-side.

**What I learned: **
Quarto has special chunk options like layout-ncol: 2 that control how figures are displayed. Using this option allows direct visual comparison between the original and transformed data by placing them side-by-side instead of stacked vertically.

::: callout-tip
1. Set the y-axis limits for both plots to go from 0 to 4000 to allow for direct comparison across plots.

2. Pay attention to `binwidth`!

3. Use a Quarto code chunk option to put the plots side-by-side.
:::