---
title: "Lab 4 Challenge"
author: "Marietta Nikolskaya"
date: "10 October, 2025"
format: 
  html:
    code-fold: true
    code-tools: true
    warning: false
    embed-resources: true
    toc: true
    toc-depth: 2
editor: source
---

# Research Question: 

#### Which California regions show the largest cost differences between center-based and family-based childcare for infants in 2018? 

```{r}
#| label: packages
library(readxl)
library(tidyverse) 
library(forcats)
```

```{r}
#| label: load-data

childcare_costs <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-05-09/childcare_costs.csv')


counties <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-05-09/counties.csv')


tax_rev <- read_csv('https://atheobold.github.io/groupworthy-data-science/labs/instructions/data/ca_tax_revenue.csv')
```


```{r}
#| label: ca-childcare-costs
ca_childcare <- childcare_costs |>
  left_join(counties, by = "county_fips_code") |>
  filter(state_abbreviation == "CA")
ca_childcare

```

```{r}
#| label: add-tax-information
ca_childcare <- ca_childcare |> 
  left_join(tax_rev, by = c(
    "county_name" = "entity_name", 
    "study_year" = "year"
  )) 
ca_childcare


```

```{r}
#| label: defining-census-regions

superior_counties <- c("Butte","Colusa","El Dorado",
                       "Glenn","Lassen","Modoc",
                       "Nevada","Placer","Plumas",
                       "Sacramento","Shasta","Sierra","Siskiyou",
                       "Sutter","Tehama","Yolo","Yuba")

north_coast_counties <- c("Del Norte","Humboldt","Lake",
                          "Mendocino","Napa","Sonoma","Trinity")

san_fran_counties <- c("Alameda","Contra Costa","Marin",
                       "San Francisco","San Mateo","Santa Clara",
                       "Solano")

n_san_joaquin_counties <- c("Alpine","Amador","Calaveras","Madera",
                            "Mariposa","Merced","Mono","San Joaquin",
                            "Stanislaus","Tuolumne")

central_coast_counties <- c("Monterey","San Benito","San Luis Obispo",
                            "Santa Barbara","Santa Cruz","Ventura")

s_san_joaquin_counties <- c("Fresno","Inyo","Kern","Kings","Tulare")

inland_counties <- c("Riverside","San Bernardino")

la_county <- "Los Angeles"

orange_county  <- "Orange"

san_diego_imperial_counties <- c("Imperial","San Diego")

ca_childcare <- ca_childcare |>
  mutate(
    county_name_clean = str_remove(county_name, " County"),
    county_name_clean = as.factor(county_name_clean),
    census_region = fct_collapse(county_name_clean,
      "Superior" = superior_counties,
      "North Coast" = north_coast_counties,
      "San Fran" = san_fran_counties,
      "N San Joaquin" = n_san_joaquin_counties,
      "Central Coast" = central_coast_counties,
      "S San Joaquin" = s_san_joaquin_counties,
      "Inland" = inland_counties,
      "LA" = la_county,
      "Orange" = orange_county,
      "San Diego" = san_diego_imperial_counties
    )
  )
ca_childcare

```

# Table 
```{r}
ca_childcare_new <- ca_childcare |> 
  filter(study_year == 2018) |>
  select(census_region, mc_infant, mfcc_infant) |> 
  mutate(cost_difference = mc_infant - mfcc_infant) |>
  group_by(census_region) |>
  summarize(
    median_center = median(mc_infant),
    median_family = median(mfcc_infant),
    median_difference = median(cost_difference)) |>
  arrange(desc(median_difference))
ca_childcare_new
```

# Scatterplot 

```{r}

scatterplot <- ca_childcare_new |>
  ggplot(aes(x = median_family, y = median_center)) + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") + 
  geom_point(color = "blue", size = 4, alpha = 0.7) +
  geom_text(aes(label = census_region), nudge_y = 10, hjust = 0) +
  labs(
    x = "Median Family Care Cost ($)", 
    y = "Median Center Care Cost ($)", 
    title = "Center vs. Family Childcare Costs for Infants by Region (2018)"
  ) + 
  theme_minimal()
scatterplot
```

# Lollipop plot 

```{r}
lollipop <- ca_childcare_new |>
  ggplot(aes(x = fct_reorder(census_region, median_difference), 
             y = median_difference)) +
  geom_segment(aes(xend = census_region, y = 0, yend = median_difference), 
               color = "gray50") +
  geom_point(size = 4, color = "steelblue") +
  coord_flip() +
  labs(
    x = "California Region",
    y = "Cost Difference: Center - Family ($)",
    title = "Cost Difference Between Center and Family Childcare for Infants (2018)"
  ) +
  theme_minimal()
lollipop
```

# Report: 

#### This analysis uses childcare cost data from California counties in 2018, examining median weekly costs for center-based versus family-based infant care across ten census regions. The research question is: Which California regions show the largest cost differences between center-based and family-based childcare for infants in 2018? Table 1 presents the median costs for both care types and their differences across all ten regions, ordered from largest to smallest difference. LA shows the largest gap at $117.94 per week, followed closely by Central Coast ($115.04) and San Francisco ($110.73). In contrast, North Coast has the smallest difference at only $47.09, roughly 2.5 times smaller than LA's gap. The scatterplot visually displays the relationship between family care costs (x-axis) and center care costs (y-axis) for each region. The dashed diagonal line represents where costs would be equal; regions positioned farther above this line have larger cost differences.This pattern suggests that higher-cost urban and coastal regions tend to have larger disparities between center-based and family-based care, potentially reflecting different market dynamics or regulatory environments in these areas. Families in regions with larger cost gaps may face more significant trade-offs when choosing between care settings.
